Cover Times of Random Walks

### 2019-05-11 paper.tex
bill's comments

For a graph with $n$ vertices, there are $n$ possible choices for the current
vertex and $2^{n-1}$ possible combinations of visited and unvisited vertices
for each one.
Of course, some cases are inaccessible by a random walk.
But even on \cref{fig:example} where we exploit symmetry,
there are almost a prohibitive number.
It is easy to see that the number of cases 
grows exponentially with the number of vertices.

@@

If we let the label of each graph in \cref{fig:example}
denote the cover time from that case, we have the following system
of equations:
\begin{align}
&\tcov = 1 + \frac{1}{3}\textrm{a} + \frac{2}{3}\textrm{b}
&\textrm{a} = 1 + \frac{1}{3}\textrm{a} + \frac{2}{3}\textrm{c} \nonumber \\
&\textrm{b} = 1 + \frac{1}{2}\textrm{b} + \frac{1}{2}\textrm{e}
&\textrm{c} = 1 + 1\textrm{e} \nonumber \\
&\textrm{d} = 1 + \frac{1}{3}\textrm{a} + \frac{1}{3}\textrm{e} + \frac{1}{3}\textrm{f}
&\textrm{e} = 1 + \frac{1}{3}\textrm{c} + \frac{1}{3}\textrm{e} + \frac{1}{3}\dot 0\nonumber \\
&\textrm{f} = 1 + \frac{1}{2}\textrm{g} + \frac{1}{2}\dot 0
&\textrm{g} = 1 + \frac{2}{3}\textrm{f} + \frac{1}{3}\dot 0\nonumber
\end{align}

With a little linear algebra, we find that $\tcov = 43/6$.

@@

In a slight abuse of notation, let $T_k$ be the 
number of steps until we reach $k$ distinct vertices.
Formally, $T_k = \min\{t: b-a+1 = k\}$.
Then by telescoping,
\begin{align}
\tcov &= \E_0[T_n] = \E[(T_2 - T_1) + (T_3 - T_2) + \cdots + 
(T_k - T_{k-1}) + \cdots + (T_n - T_{n-1})] \nonumber
\end{align}

@@

Considering the cases that we either reach $b$ or not before
returning to $a$,
\begin{align}
\E_a[T_b] + \E_b[T_a] &= 
\E_a[T_a | T_b < T_a] + 
\E_a[T_a | T_a < T_b] (\frac{1}{\pr(T_b<T_a)} - 1) \nonumber.
\end{align}
Observe that $1/\pr(T_b<T_a)-1 = (1-\pr(T_b<T_a))/\pr(T_b<T_a)$, 
which simplifies to $\pr(T_a<T_b)/\pr(T_b<T_a)$.
Multiplying by the probability we reach $b$ before returning to $a$,
\begin{align}
\E_a[T_a|T_b < T_a] \pr(T_b<T_a) + \E_a[T_a|T_a<T_b] \pr(T_a<T_b)=\frac{1}{\pi_a} \nonumber
\end{align}
and the commute identity follows
\begin{align}
E_a[T_b] + E_b[T_a] = \frac{1}{\pi_a}\frac{1}{\pr_a(T_b < T_a)} = \Reff C.
\label{eqn:commute}
\end{align}

@@

\begin{proof}
Without loss of generality, label the vertices $\{1, 2, \dots, n\}$.
We may assume that the walk started at vertex $n$.
Let $\sigma$ be a uniform permutation of the unvisited $n-1$ vertices.
The strategy is to look for the vertices in order $\sigma$.

@@

We begin by writing the cover time in a clever way
(similar to the strategy we employed on the cycle graph):
\begin{align}
\tcov &= \E_n[{T_{n-1}}]  \nonumber \\
&= \E_n[T_1 + (T_2 - T_1) + \cdots + 
(T_k - T_{k-1}) + \cdots + (T_{n-1} - T_{n-2})] . \nonumber
\end{align}

@@

The same strategy for the Matthews upper bound can be used
to find a lower bound.
Instead of looking for all $n$ vertices, we search
for $A \subseteq V(G)$.
When the hitting time is large between any two vertices in $A$,
the time to visit all of $A$ is a good lower bound on the cover time
of $V(G)$.

@@

By \cref{eqn:commute}, the commute time between the root and a leaf $a$ is
\begin{align}
C R_{\rho, a} = 2 (n-1) k \nonumber.
\end{align}
Note that effective resistance adds in series so $\Rxy = i$ where $i$
is the length of the single path between $x$ and $y$.

@@

and \cref{thm:matthews_up} gives
\begin{align}
\tcov &\leq 2(n-1)k\left(1 + \frac{1}{2} + \cdots + \frac{1}{n} \right) \nonumber \\
&\approx (2 - \frac{2}{n}) \cdot n k \log n \approx (2 - \frac{2}{n}) n k \log 2^{k+1} -1 \nonumber \\
&\approx (2  + o(1)) \cdot n k^2 \log 2 \nonumber
\end{align}

@@

In the worst case, $h' = h-1$.
Then by \cref{thm:matthews_low},
\begin{align}
\tcov &\geq 2 (n-1)(k-h+1)\left(1 + \frac{1}{2} + \cdots + \frac{1}{2^h -1} \right) \nonumber \\
&\approx (2 + o(1)) \cdot n (k-h) h \log 2 \nonumber.
\end{align}

@@

In \cref{sec:commute}, we fixed vertices $a$ and $b$ and called
the effective resistance between them $\Reff$.
We now consider $\Rxy$ the effective resistance between vertices $x$ and $y$.

@@

The expected time of this particular path on the original graph is certainly less than
or equal to the cover time. That is,
\begin{align}
\tcov &\leq \sum_{i=0}^{2n-3} \E_{v_i}[T_{v_{i+1}}]. \nonumber
\end{align}

@@

We compute the variance of $N_n$ by finding $\E[N_n^2]$ and $\E[N_n]^2$.
Let $\I_m$ be the indicator random variable for event $A_m$.
That is, $\I_m = 1$ if vertex $m$ is unvisited 0 otherwise.
We already know $\E[N_n]$ so we want to find $\E[N_n^2]$,
\begin{align}
\E[N_n^2] &= \E \left[ \left(\sum_{m=1}^n \I_m \right) ^2 \right] =
\E \left[ \sum_{m=1}^n \sum_{k=1}^n \I_m \I_k \right] =
\sum_{1 \leq k, m \leq n} \pr(A_k \cap A_m). \nonumber
\end{align}

The second equality holds since we distribute terms.
The third equality holds since the mean of an indicator for event $A_m$
is the probability of $A_m$.
Putting both terms together, we find the variance
\begin{align}
\Var[N_n] &= \E[N_n^2] - \E[N_n]^2 \nonumber \\
&= \sum_{1 \leq k, m \leq n} \pr(A_k \cap A_m) - P(A_k)P(A_m) \nonumber \\
&= n(n-1) \{ \pr(A_k \cap A_m) - P(A_k)P(A_m) \} + 
n \{ \pr(A_k \cap A_m) - P(A_k)P(A_m) \} \nonumber \\
&= n(n-1) \{ (1-2/n)^r - (1-1/n)^{2r} \} + 
n \{ (1-1/n)^r  - (1-1/n)^{2r} \}. \nonumber
\end{align}
where $k\neq m$ in the first term and $k=m$ in the second.

@@

Thus we conclude that $N_n/n \rightarrow \E[N_n/n] = e^{-c}$ in probability.

@@

We use \cref{thm:pois} to find the distribution of the cover time variable $\cov$.
Set $r = n \log n + nx$.
Then 
\begin{align}
ne^{-r/n} &= n e ^{- \log n - x} \nonumber \\
&= n e^{- \log n} e^ {- x} = e^{-x}.\nonumber
\end{align}

Using the fact that $\cov \leq m$ if and only if the first $m$ steps
of the random walk visit all $n$ vertices
and $ne^{-r/n} \rightarrow e^{-x}$ as the assumption of \cref{thm:pois},
\begin{align}
\pr(\cov \leq r) &\rightarrow e^{-e^{-x}} \nonumber \\
\pr(\cov - n \log n\leq nx) &\rightarrow e^{-e^{-x}}. \label{eqn:dist}
\end{align}

>>>>

For a graph with $n$ vertices, there are $n$ possible choices for the current
vertex and $2^{n-1}$ possible combinations of visited and unvisited vertices
for each one.
Of course, some cases are inaccessible by a random walk.
But even on \cref{fig:example} where we exploit symmetry,
there are almost a prohibitive number.
It is easy to see that the number of cases 
grows exponentially with the number of vertices.
\begin{comment}
\begin{figure}[ht]
	\centering
		\begin{tikzpicture} [
            > = stealth, % arrow head style
            shorten > = 1pt, % don't touch arrow head to node
            %auto,
            node distance = 3cm, % distance between nodes
            semithick % line style
        ]

@@

			\node (A) at (-1.2,2.5) {
				    \node (l) at (.5, 1.1) {A};
				    \node (tl) at (0,1) {};
				    \node[current] (tr) at (1,1) {};
			\node (B) at (2,3.5) {
				    \node (l) at (.5, 1.3) {B};
			\node (C) at (2,1) {
				    \node (l) at (.5, -.3) {C};
			\node (D) at (4,3.5) {
				    \node (l) at (.5, 1.3) {D};
			\node (E) at (4,1) {
				    \node (l) at (.5, -.3) {E};
			\node (F) at (6,3.5) {
				    \node (l) at (.5, 1.3) {F};
			\node (G) at (6, 1) {
				    \node (l) at (.5, -.3) {G};
			\node (H) at (9.2, 2.5) {
				    \node (l) at (.5, 1.1) {H};
			\path[->] (A) edge (B);
			\path[<->] (B) edge (C);
			\path[->] (B) edge (D);
			\path[->] (B) edge (E);
			\path[->] (C) edge (E);
			\path[<->] (D) edge (F);
			\path[->] (D) edge[bend left, distance=12mm] (H);
			\path[<->] (E) edge (G);
			\path[->] (E) edge[bend right, distance=15mm] (H);
			\path[->] (E) edge[loop above, distance=5mm] (E);
			\path[->] (F) edge (H);
			\path[->] (G) edge (H);

If we let the label of each graph in \cref{fig:example}
denote the cover time from that case, we have the following system
of equations:
\begin{comment}
\begin{align}
&\tcov = 1 + \frac{1}{3}\textrm{a} + \frac{2}{3}\textrm{b}
&\textrm{a} = 1 + \frac{1}{3}\textrm{a} + \frac{2}{3}\textrm{c} \nonumber \\
&\textrm{b} = 1 + \frac{1}{2}\textrm{b} + \frac{1}{2}\textrm{e}
&\textrm{c} = 1 + 1\textrm{e} \nonumber \\
&\textrm{d} = 1 + \frac{1}{3}\textrm{a} + \frac{1}{3}\textrm{e} + \frac{1}{3}\textrm{f}
&\textrm{e} = 1 + \frac{1}{3}\textrm{c} + \frac{1}{3}\textrm{e} + \frac{1}{3}\dot 0\nonumber \\
&\textrm{f} = 1 + \frac{1}{2}\textrm{g} + \frac{1}{2}\dot 0
&\textrm{g} = 1 + \frac{2}{3}\textrm{f} + \frac{1}{3}\dot 0\nonumber
\end{align}
\end{comment}
\begin{align}
\tcov = &\textrm{A} = 1 + \textrm{B}
&\textrm{B} = 1 + \frac{1}{2}\textrm{C} + \frac{1}{2}\textrm{D} \nonumber \\
&\textrm{C} = 1 + \frac{1}{2}\textrm{B} + \frac{1}{2}\textrm{E}
&\textrm{D} = 1 + \frac{1}{2}\textrm{F} + \frac{1}{2}\textrm{H} \nonumber \\
&\textrm{E} = 1 + \frac{1}{3}\textrm{E} + \frac{1}{3}\textrm{G} + \frac{1}{3}\textrm{H}
&\textrm{F} = 1 + \frac{1}{2}\textrm{D} + \frac{1}{2}\textrm{F} \nonumber \\
&\textrm{G} = 1 + \frac{1}{2}\textrm{E} + \frac{1}{2}\textrm{H}
&\textrm{H} = 0\nonumber
\end{align}

With a little linear algebra, we find that $\tcov = 43/5$.
We leave it as an exercise to the reader to verify that the time
to visit all vertices from the other starting vertex is $43/6$.

@@

In a slight abuse of notation, let $T_k$ be the 
number of steps until we reach $k$ distinct vertices.
Formally, $T_k = \min\{t: b-a+1 = k\}$.
Then by telescoping,
\begin{align}
\tcov &= \E[T_n] = \E[(T_2 - T_1) + (T_3 - T_2) + \cdots + 
(T_k - T_{k-1}) + \cdots + (T_n - T_{n-1})] \nonumber
\end{align}

@@

The expected commute time between $a$ and $b$ is the
expected return time to $a$ given we reach $b$ plus
the expected return time to $a$ given we do not reach $b$ times
the number of times we return to $a$ before the success when
we reach $b$.
Formally, we write
\begin{align}
\E_a[T_b] + \E_b[T_a] &= 
\E_a[T_a | T_b < T_a] + 
\E_a[T_a | T_a < T_b] (\frac{1}{\pr(T_b<T_a)} - 1) \nonumber.
\end{align}
Observe that $1/\pr(T_b<T_a)-1 = (1-\pr(T_b<T_a))/\pr(T_b<T_a)$, 
which simplifies to $\pr(T_a<T_b)/\pr(T_b<T_a)$.
Multiplying by the probability we reach $b$ before returning to $a$,
\begin{align}
\E_a[T_a|T_b < T_a] \pr(T_b<T_a) + \E_a[T_a|T_a<T_b] \pr(T_a<T_b)=\frac{1}{\pi_a} \nonumber
\end{align}
and the commute identity follows
\begin{align}
E_a[T_b] + E_b[T_a] = \frac{1}{\pi_a}\frac{1}{\pr_a(T_b < T_a)} = \Reff C.
\label{eqn:commute}
\end{align}

@@

\begin{proof}
Without loss of generality, label the vertices $\{1, 2, \dots, n\}$
such that $n$ is the vertex from which it takes the most steps to cover
the remaining vertices.
Let $\sigma$ be a uniform permutation of the unvisited $n-1$ vertices.
The strategy is to look for the vertices in order $\sigma$.

@@

We begin by writing the cover time in a clever way
(similar to the strategy we employed on the cycle graph):
\begin{align}
\tcov &= \E_n[{T_{n-1}}]  \nonumber \\
&\leq \E_n[T_1 + (T_2 - T_1) + \cdots + 
(T_k - T_{k-1}) + \cdots + (T_{n-1} - T_{n-2})] . \nonumber
\end{align}

@@

The same strategy for the Matthews upper bound can be used
to find a lower bound.
Instead of looking for all $n$ vertices, we search
for $A \subseteq V(G)$.
The Matthews lower bound will always hold.
However, we can tighten the bound by picking $A$ such
that the hitting time is large between any pair of vertices.
We use large in a relative sense and 
offer an illustrative example in \cref{sec:binary}.

@@

By \cref{eqn:commute}, the commute time between the root and a leaf $a$ is
\begin{align}
C R_{\rho, a} = 2 (n-1) k \nonumber.
\end{align}
Note that resistance adds in series so 
effective resistance is just the sum of the resistances along a path.
Then $\Rxy = i$ where $i$
is the length of the single path between $x$ and $y$.

@@

and \cref{thm:matthews_up} gives
\begin{align}
\tcov &\leq 2(n-1)k\left(1 + \frac{1}{2} + \cdots + \frac{1}{n} \right) \nonumber \\
&\sim (2 - \frac{2}{n}) \cdot n k \log n 
\sim(2 - \frac{2}{n}) n k \log 2^{k+1} -1 \nonumber \\
&\sim(2  + o(1)) \cdot n k^2 \log 2 \nonumber
\end{align}

@@

In the worst case, $h' = h-1$.
Then by \cref{thm:matthews_low},
\begin{align}
\tcov &\geq 2 (n-1)(k-h+1)\left(1 + \frac{1}{2} + \cdots + \frac{1}{2^h -1} \right) \nonumber \\
&\geq (2 + o(1)) \cdot n (k-h) h \log 2 \nonumber.
\end{align}

@@

In \cref{sec:commute}, we fixed vertices $a$ and $b$ and called
the effective resistance between them $\Reff$.
We now consider $\Rxy$ the effective resistance between vertices $x$ and $y$.
Recall that $C$ is the sum of the conductances of every vertex in our graph.

@@

The expected time of this particular path on the original graph is certainly 
greater than or equal to the cover time. That is,
\begin{align}
\tcov &\leq \sum_{i=0}^{2n-3} \E_{v_i}[T_{v_{i+1}}]. \nonumber
\end{align}

@@

We compute the variance of $N_n$ by finding $\E[N_n^2]$ and $\E[N_n]^2$.
Let $\I_m$ be the indicator random variable for event $A_m$.
That is, $\I_m = 1$ if vertex $m$ is unvisited 0 otherwise.
We first find $\E[N_n^2]$
\begin{align}
\E[N_n^2] &= \E \left[ \left(\sum_{m=1}^n \I_m \right) ^2 \right] =
\E \left[ \sum_{m=1}^n \sum_{k=1}^n \I_m \I_k \right] =
\sum_{1 \leq k, m \leq n} \pr(A_k \cap A_m). \nonumber
\end{align}
The second equality holds since we distribute terms.
The third equality holds since the mean of an indicator for event $A_m$
is the probability of $A_m$ and both events must occur for
us to have a nonzero term.
Then we find $\E[N_n]^2$
\begin{align}
\E[N_n]^2 &= \E \left[ \left(\sum_{m=1}^n \I_m \right)
\left( \sum_{k=1}^n \I_k \right) \right] =
\sum_{1 \leq k, m \leq n} \pr(A_k) \pr(A_m).
\nonumber
\end{align}

Putting both together, we find the variance
\begin{align}
\Var[N_n] &= \E[N_n^2] - \E[N_n]^2 \nonumber \\
&= \sum_{1 \leq k, m \leq n} \pr(A_k \cap A_m) - P(A_k)P(A_m) \nonumber \\
&= n(n-1) \{ \pr(A_k \cap A_m) - P(A_k)P(A_m) \} + 
n \{ \pr(A_k \cap A_m) - P(A_k)P(A_m) \} \nonumber \\
&= n(n-1) \{ (1-2/n)^r - (1-1/n)^{2r} \} + 
n \{ (1-1/n)^r  - (1-1/n)^{2r} \}. \nonumber
\end{align}
where $k\neq m$ in the first term and $k=m$ in the second.

@@

Thus we conclude that $N_n/n \rightarrow e^{-c}$.

@@

We use \cref{thm:pois} to find the distribution of the cover time variable $\cov$.
Set $r = n \log n + nx$.
Then 
\begin{align}
ne^{-r/n} &= n e ^{- \log n - x} = n n ^{-1} e^ {- x} = e^{-x}.\nonumber
\end{align}

Using the fact that $\cov \leq m$ if and only if the first $m$ steps
of the random walk visit all $n$ vertices
and $ne^{-r/n} \rightarrow e^{-x}$ as the assumption of \cref{thm:pois},
\begin{align}
P(N_n=0) = \pr(\cov \leq r) = \pr(\cov - n \log n\leq nx) 
&\rightarrow e^{-e^{-x}}. \label{eqn:dist}
\end{align}
